---
title: "Gov 50 Final Project"
author: "Annika Gagnon"
description: "akgagnon Final Project"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
#Options
options(
  scipen = 9,
  width = 100)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center")

library(tidyverse)
library(broom)

full <- read_csv("data/mentalhealth.csv")
names(full) <- gsub(" ", "_", names(full)) |> tolower()

mental <- full |>
  filter(group == "By Sex",
         indicator == "Symptoms of Anxiety Disorder or Depressive Disorder")
```

# Introduction

The U.S. Census Bureau created the Household Pulse Survey to track the effects of the Covid-19 pandemic on American households. The survey was made to measure the impact of the pandemic on employment, consumer spending, food security, housing, disruptions in educational careers, and physical and mental well-being. For the purposes of this study, I will be focusing on how the pandemic affected mental health. Using this data, I will investigate how reported mental health changes over the course of the survey and how it differs between men and women.

# Data

The Household Pulse Survey was conducted from April of 2020 through October of 2022. The survey was sent out via email and text message and was conducted online. Households that were linked to one or more email addresses or phone numbers were selected randomly to participate. Surveys asking about the mental health of participants were sent out 50 times over the course of the the study. Data from these surveys was collected during periods ranging from 5 days to two weeks. The mental health of survey subjects was measured using a four point scale. The percentage of subjects who reported feelings symptoms of anxiety and/or depression was then recorded. It is the percentage numbers that I will be working with in this study; thus, when "points" are discussed, assume that we are working off of a 100 point scale.

I hypothesize that over the course of the pandemic, the percentage of people experiencing symptoms of anxiety and/or depression will increase due to increased isolation. 

According to a study done pre-pandemic by the American Psychological Association, women are more likely to be diagnosed with anxiety and/or depression. Because of this, I hypothesize that the percentage of women who experience symptoms of anxiety and/or depression will be greater than the percentage of men who experience symptoms of anxiety and/or depression throughout the course of the whole study. 

This analysis covers responses from over 2 million participants over the course of 50 survey time periods. My research design is observational, correlational analysis and makes no causal assumptions or claims.


```{r}
mental|>
  ggplot(aes(x = value)) +
  geom_histogram() +
  labs(x = "Percentage of People Reporting Symptoms of Anxiety and/or Depression", y = "Frequency")
```
The results of the Household Pulse Survey provide us with the outcome variable: the percentage of people reporting symptoms of anxiety and/or depression during a given time period. The outcome ranges from 24.4 to 47.6 with a median value of 34.75 and a mean of 34.99. This means that over the course of the 50 data collection time periods, results ranged from 24.4% to 47.6% of survey participants reporting symptoms of anxiety and/or depression. On average, 34.99% of survey participants reported feeling symptoms of anxiety and/or depression.

# Results

To compare the the amount of women reporting symptoms to the amount of men reporting symptoms, consider the graph below.

```{r}
ggplot(data = mental,
       mapping = aes(x = subgroup,
                     y = value)) +
  geom_boxplot() +
  labs(x = "Gender",
       y = "Percentage Reporting Symptoms of Anxiety and/or Depression")
  
```
The above bar-plot suggests that on average, 38.69% of women surveyed reported experiencing symptoms of anxiety and/or depression while 31.29% of men surveyed reported experiencing symptoms of anxiety and/or depression.

```{r}
mental|>
    group_by(time_period, subgroup) |>
    summarize(average = mean(value, na.rm = TRUE)) |>
    ggplot(aes(x = time_period, y = average)) +
    geom_line(aes(col = subgroup)) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Time Period", y = "Percentage of People Reporting Symptoms of Anxiety and/or Depression", title = "Symptoms of Anxiety and/or Depression in Men vs Women", subtitle = "Over the course of the Covid-19 Pandemic")

#consider using a bar plot median mental health men vs women
```
The above plot helps to better visualize the data so that the third variable time is included. The plot suggests that there is a downward correlation between the amount of people reporting experiencing symptoms of anxiety and/or depression regardless of gender. It also suggests that women were consistently more likely to report experiencing symptoms of anxiety and/or depression over the course of the 50 survey time periods. Regression analysis will provide a more in depth look into this seemingly negative relationship.


# Regression Results

The following tables are the results from two regression models. The first is a bivariate model which regresses the percentage of people reporting symptoms of anxiety and/or depression on time period. The second is a multivariate model which takes into account the gender of the people reporting symptoms of anxiety and/or depression. 

```{r}
bi_lm <- lm(
  data = mental,
  value ~ time_period)

modelsummary::modelsummary(bi_lm,
                           title = "Bivariate Model",
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

The bivariate model suggests that on average and holding all else constant, all people participating in the study regardless of gender reported experiencing less symptoms of anxiety and/or depression from the time the survey began to the time that it ended. This is determined by the value of the coefficient "time_period". The value is -0.139 which tells us that on average for every one time-period increase, the percentage of people reporting symptoms of anxiety and/or depression decreased by 0.139 points.

```{r}
multi_lm <- lm(
    data = mental|>
      mutate(female = if_else(subgroup == "Female", 1,0)),
      value ~ female + time_period)

modelsummary::modelsummary(multi_lm,
                           title = "Multivariate Model",
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

The multivariate model suggests that on average and holding all else constant, women participating in the study reported experiencing less symptoms of anxiety and/or depression from the time the survey began to the time that it ended. This is determined by the value of the coefficient "time_period". The value is -0.139 which tells us that on average for every one time-period increase, the percentage of women reporting symptoms of anxiety and/or depression decreased by 0.139 points. 

The model also shows us that the percentage of women reporting symptoms of anxiety and/or depression was 7.41 points higher than the percentage of men reporting symptoms of anxiety and/or depression. On a 100 point scale this may not seem like much, but considering that results range from 24.4 to 47.6 which is a difference of only 23.2 points, it may be more significant than it appears at first glance.

# Statistical significance

In both the bivariate and multivariate models, the association between reported symptoms of anxiety and/or depression and time period is statistically significant at the test level α = 0.05.The p-value on the coefficient in both models is < 0.001. If hypothesizing that the relationship between symptoms of anxiety and/or depression and time period was non-zero, then the coefficient’s p-value of > 0.001 would suggest there is a near 0% probability that the estimated association observed here is as insignificant as any estimated relationship from the null distribution. This means that there is a nearly 100% probability that the estimated relationship is not a random result but rather a statistically significant one because it is so different from what the relationship would look like in the null distribution.

The R^2 value of the bivariate model is 0.141 and the R^2 value of the multivariate model is 0.619. This means that the data does not fit the bivariate model very well, but it fits the multivariate model better. I suspect this is because the multivariate model is dealing with a single gender instead of two. I am inclined to think that if the regression for the multivariate model was repeated with men instead of women, the data would fit the model similarly well. 

If we assume that the sampling distribution of survey time periods is approximately normal, then the standard error on the coefficient estimates for reported symptoms of anxiety and/or depression can be used to construct a confidence interval of 95%. The z-score that captures 95% of the data assuming a normal distribution is 1.96. 1.96 can therefore be multiplied by the coefficient standard error and that product can be added and subtracted to the coefficient point estimate to produce the upper and lower bound limits of a 95% confidence interval.

The bivariate model 95% confidence interval is below.

βPreferences = −0.139∈(-0.2082754, -0.07040409)

The multivariate model 95% confidence interval is below.

βPreferences = −0.139∈(-0.1854745, -0.09320495)

This means that 95% of regression estimated correlations between the percentage of people reporting symptoms of anxiety and/or depression and time period would be within the intervals listed above. On a deeper level, this means that 95% of the results depict a weak negative correlation between symptoms of anxiety and/or depression and time period meaning that over the course of the pandemic, people's reported mental health improved.

# Causality and Confounders

The explanatory variable time period is not randomly assigned and thus cannot be viewed as a "treatment" variable. Therefore, change in time period cannot be said to causally increase or decrease the percentage of people surveyed who report experiencing symptoms of anxiety and/or depression. Time period and reported mental health are negatively correlated and this correlation is statistically significant. 

There are several opportunities for errors and confounding to occur within this study. 

Measurement errors can occur if those being surveyed provide incorrect information due to an unclear survey question or misunderstanding. Furthermore, the Pulse Survey was only available for limited time periods so survey subjects were working against a time limit.

Coverage errors could have occurred. This means that there are millions of individuals who could have been included in the survey and were not. The survey was only able to send out surveys to households that had identifiable email addresses or phone numbers attached to them.

This lack of coverage provides us with a confounder: lack of internet or cell phone access. Because the Pulse Survey was conducted online or via cell phone, American households that do not have access to the internet or a cell phone are unable to participate. 

Other confounders such as household income, the political climate of the state one lives in, and access to clean water and food also exist but are excluded from this study due to lack of data. Lack of financial and natural resources can put a strain on the mental health of households. Lock-downs were more common in politically progressive states than politically conservative states; thus, the political climate of the state one lives in plays a large role in determining how isolated a household is. Not having access to clean water and food is always stressful, but for some American households lack of access to these key resources became exacerbated by the pandemic.


# Conclusion

The results of the regressions demonstrate that as time went on during the first two years of the Covid-19 pandemic, people began to report experiencing less symptoms of anxiety and/or depression. The multivariate model demonstrates that women on average reported experiencing more symptoms of anxiety and/or depression than men did. 

This study is limited in regard to the type of data collected. Because the only variables being studied are gender, reported mental health, and time, it is difficult to gage what about the pandemic affected the reported mental health of American households.

A second limitation is the amount of data used for this study. To gain a better understanding of how the Covid-19 pandemic affected the mental health of the nation it would be better to look at data from before the pandemic and continue the study beyond present day as Covid mitigation in the U.S. becomes more relaxed.

# Next Steps

This study is very simple and I do not think that it does a great job of exploring how the pandemic has effected mental health. If I were to repeat this study, I would look into states in the US that went into lock-down during the pandemic. I think it would be interesting to look into how those surveyed would report their mental health during survey time periods while their state was in lock-down vs time periods when their state was not in lock-down.

In terms of improving this current study, I think it would be helpful to combine the bivariate and multivariate regression models into one model so that the values can be compared side by side. I think that this adjustment would provide readers of this study with more clarity on the outcomes of the regressions.



# Sources

https://www.apa.org/news/press/releases/2011/08/mental-illness





